Black swans in ecological time series\\ Black swans in population dynamics\\ Heavy-tailed ecological time series\\ Black swans in ecological dynamics\\ Black swan ecological dynamics\\ \ldots

### 2014-10-29 text/ms.tex
Edit introduction

Black swans are unexpected extreme events with potentially dramatic
consequences \citep{taleb2007,sornette2009}. One of the most striking black
swans in ecology was the asteroid marking the mass extinction of the K-T
boundary \citep{harnik2012}. Today, it is the extremeness of climate, in
concert with shifts in mean temperature, that is expected to cause the greatest
societal damage \citep{ipcc2012}. But, while extremes in the physical
environment such as wave height, rain fall, and temperature are present and
widely accepted \citep{gaines1993,katz2005}, it remains unclear the extent to
which ecological systems buffer or suffer from black swans \citep{nunez2012}.

%Existing evidence for black swans or heavy tails in population dynamics is
%mostly anecdotal or for particular systems.
There is compelling anecdotal evidence for ecological black swans, but
systematic evidence has been elusive. A survey of ecologists indicated that
surprising results in field experiments may be far more common than we assume
\citep{doak2008}, and a number of major environmental events on geologic and
contemporary time scales could be considered to have caused ecological black
swans \citep{nunez2012}. Other work has detected heavy-tail dynamics for
particular systems. For example, \citet{segura2013} identified heavy tails in
time series of marine microbe populations and \citet{drake2014} identified
heavy tails in time to extinction for experimental waterflea populations.
However, systematic evidence has been limited. What appeared to be heavy tails
in breeding bird abundance \citep{keitt1998}, could also be explained by the
mixing of species in the original dataset \citep{allen2001}. A subsequent
large scale analysis of the Global Population Dynamics Database (GPDD) found
little evidence for heavy-tail distributions in longer abundance time series
\citet{halley2002}.

There are two key reasons why we may find little evidence of ecological black
swans. First, they might not systematically exist. Indeed, the majority of
model fitting and risk forecasting assumes that population dynamics are not
heavy tailed \citep[e.g.][]{brook2006a,dennis2006,knape2012}. Alternatively,
dynamics might be heavy tailed, but a paucity of statistical methods
appropriate for commonly available series has limited our ability to detect
them. For example: (1) The generalized extreme value (GEV) distribution has
been applied to environmental data \citep[e.g.][]{katz2005}. The GEV describes
the distribution of the most extreme event per time interval (e.g.~heaviest
rainfall per year), but this approach requires long time series that can be
condensed into time intervals. (2) Thin- and heavy-tailed distributions have
been competed with each other in an information theoretic framework
\citep{halley2002}, but this analysis does not quantify the probability of
black swans on a continuous scale or allow for population dynamics. (3) Fitting
a catastrophic mixture distribution state space model allows one to quantify
the probability that population events are catastrophic but is also data
intensive \citep{ward2007}.

>>>>

Black swans are unexpected extreme events with potentially dramatic
consequences \citep{taleb2007,sornette2009}. One of the most striking black
swans in ecology was the asteroid marking the mass extinction of the K-T
boundary \citep{harnik2012}. Today, it is the extremeness of climate, in
concert with shifts in mean temperature, that is expected to cause the greatest
societal damage \citep{ipcc2012}. But, while extremes in the physical
environment such as wave height, storm severity, and temperature are present and
widely accepted \citep{gaines1993,katz2005}, it remains unclear the extent to
which ecological systems buffer or suffer from black swans \citep{nunez2012}.

%Existing evidence for black swans or heavy tails in population dynamics is
%mostly anecdotal or for particular systems.
There is compelling anecdotal evidence for ecological black swans, but
systematic evidence has been elusive. A survey of ecologists indicated that
surprising results in field experiments may be far more common than we assume
\citep{doak2008}, and a number of major environmental events on geologic and
contemporary time scales could be considered to have caused ecological black
swans \citep{nunez2012}. Other work has detected heavy-tail dynamics for
particular systems. For example, \citet{segura2013} identified heavy tails in
time series of marine microbe populations and \citet{drake2014} identified
heavy tails in time to extinction for experimental waterflea populations.
However, systematic evidence has been limited. What appeared to be heavy tails
in breeding bird abundance \citep{keitt1998}, could also be explained by 
mixing of species in the original dataset \citep{allen2001}. A subsequent
large scale analysis of the Global Population Dynamics Database (GPDD) found
little evidence for heavy-tail distributions in longer abundance time series
\citet{halley2002}.

There are two key reasons why we may find little evidence of ecological black
swans. First, they might not systematically exist. Indeed, the majority of
model fitting and risk forecasting assumes that population dynamics are not
heavy tailed \citep[e.g.][]{brook2006a,dennis2006,knape2012}. Alternatively,
dynamics might be heavy tailed, but a paucity of statistical methods
appropriate for commonly available series has limited our ability to detect
them. For example: (1) The generalized extreme value (GEV) distribution has
been applied to environmental data \citep[e.g.][]{katz2005}. The GEV describes
the distribution of the most extreme event per time interval (e.g.~heaviest
rainfall per year), but this approach requires long time series that can be
condensed into time intervals. (2) Thin- and heavy-tailed distributions have
been competed with each other in an information theoretic framework
\citep{halley2002}, but this analysis did not quantify the probability of
black swans on a continuous scale or allow for population dynamics. (3) Fitting
a catastrophic mixture distribution state space model allows one to quantify
the probability that population events are extreme on a continuous scale
but is data intensive \citep{ward2007}.
